---
title: "Education_employment_causal_inference_analysis"
author: "StudentA"
date: "12/9/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Abstract
Education believes to have substantial effect on employment prospects. This paper aims to investigate the causal links between education and employment by employing propensity score method. Demonstration of the propensity score matching method is presented in examining the causal effect of education on employment. 

## Keywords
propensity scores, causal inference, employment, education

## Introduction
Statistical analysis has been widely implemented in carrying out researches. It involves collecting and selecting data, drawing meaningful conclusions, and reporting findings. In drawing conclusions about a causal link based on varying conditions of an effect, causal inferences is the crucial process. Causal inference analyzes the response of the effect variable with cause being changed. Observational data is often considered more feasible and reliable than experimental data. It is thus of great importance to use causal inference in investigating potential relationship between two variables.

One problem present in observational data is confounding. It impedes one's ability to draw causal inferences. Propensity score matching methods is thus introduced and has been widely implemented then. One of the propensity method, namely matching is presented in this context of examination of causal effect of education on employment rate. I will implement this method to discern if there is a causal link between whether or not a person receives high education and whether or not a person is employed.

Dataset obtained from Voter study group will be used. Propensity score matching method is implemented in making inference on the causal link between education and employment. In the Methodology section, data and the model used for propensity analysis are described. Results section includes the result derived from the propensity score analysis from the previous section. Inference drawn on the data and conclusions are included in the final Conclusion section.

## Methodology
# Data 
github repo link: https://github.com/js25-lab/Education_employment_causal_inference_Analysis/blob/main/Causal_analysis/Education_employment_causal_relationship_analysis.Rmd
```{r}
# Setting up workspace
library(tidyverse)
# Read raw data
raw_data <- read_dta("ns20200625.dta")
# Adding labels
raw_data <- labelled::to_factor(raw_data)
raw_data
# Kept variables 
reduced_data <-
  raw_data %>% 
  select(employment,
         gender,
         race_ethnicity,
         education,
         state,
         household_income,
         age)
```
# Model




## Results




## Discussion
# Summary
# Conclusions
# Weaknesses & Next Steps



## Appendices



## References